Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-07-03 docs/paper.Rmd
write data & empirical strategy

>>>>

\section{Data and Empirical Strategy}

The analysis draws on multiple nationally representative data sources, with the Philippine Censuses serving as the primary dataset. Microdata from the 1990, 1995, 2000, 2007, and 2010 waves are used to construct baseline comparisons and perform robustness checks. The 1995 wave is treated as the pre-policy reference period, while 2007 serves as the post-policy period. Due to the possibility that reductions in contraceptive access began prior to the formal issuance of Executive Order No. 003 in 2000, the 2000 census wave is excluded from the baseline specification, though it is incorporated in secondary analyses. Placebo estimations using the 1990 census are also implemented to validate the identifying assumptions.

While the census provides rich information on household composition and allows for city-level fixed effects with considerable statistical power, it does not include direct measures of child health or anthropometry. To address this limitation, two additional datasets are employed. The Philippine Demographic and Health Surveys (DHS), pooled across the 1993, 1998, 2003, and 2008 rounds, offer standardized anthropometric data for children under five, alongside detailed maternal characteristics and fertility histories. The DHS enables construction of child-level outcomes—height-for-age z-scores (HAZ) and weight-for-height z-scores (WHZ)—and supports temporal comparisons across the pre- and post-ban periods. These outcomes serve as internationally recognized indicators of chronic and acute malnutrition, respectively.

To complement these sources, the Annual Poverty Indicators Survey (APIS) is used to assess household-level economic conditions, particularly food expenditure, living standards, and maternal education. Although smaller in scale, the APIS provides contextual variables necessary for exploring heterogeneity in treatment effects across wealth strata or educational backgrounds.

The empirical strategy adopts a difference-in-differences (DiD) framework. It utilizes the 2000 contraceptive policy in Manila as a natural experiment. The policy—enacted through Executive Order No. 003—effectively banned modern contraceptive distribution in public health facilities across Manila City, but left surrounding municipalities in the National Capital Region (NCR) unaffected. This jurisdictional variation enables identification of the causal effect of fertility increases on child nutrition, under the assumption that, in the absence of the policy, nutritional trends in Manila would have followed a parallel trajectory to those in comparison cities.

The main estimating equation is specified as follows:

\[
\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict},
\]

where \(\text{Nutrition}_{ict}\) denotes the anthropometric outcome (HAZ or WHZ) for child \( i \) in city \( c \) at time \( t \); \(\text{Post}_t\) is an indicator for the post-policy period; \(\text{Manila}_c\) is a treatment group dummy; \(X_{ict}\) is a vector of controls including child age, sex, birth order, maternal age and education, household wealth quintile, and urban barangay indicators; \(\lambda_c\) and \(\tau_t\) are city and year fixed effects, respectively. The coefficient \(\beta\) estimates the average treatment effect of the fertility shock on nutritional outcomes.

To strengthen identification, additional specifications are estimated. City-specific linear trends are introduced to account for differential temporal patterns in nutrition outcomes. A triple-difference extension interacts the DiD term with maternal age cohort indicators, under the assumption that older women—being less fecund—were less likely to experience changes in fertility behavior. This approach allows for control of unobserved city-level shocks that may disproportionately affect certain demographic groups. Sensitivity analyses also test alternative control groups: first, using only Quezon City, which is geographically contiguous and socio-economically comparable to Manila; and second, using all other NCR cities excluding Quezon.

Standard errors are clustered at the city-year level to allow for serial correlation and heteroskedasticity. Placebo tests using the 1990–1995 period confirm that pre-policy trends in nutritional outcomes were parallel across treatment and control groups. Additional robustness checks include falsification tests using children born outside of the city jurisdiction and exclusion of households with missing maternal linkage or implausible anthropometric values.

Although the analysis is subject to the usual limitations of observational designs—including residual confounding and measurement error—the combined use of census, DHS, and APIS data enables a multi-faceted examination of the Q–Q trade-off in nutrition. The strategy mirrors that of \citet{dumas2019sex}, who investigated educational outcomes using the same policy shock, but extends their framework to explore a dimension of child quality that is both biologically foundational and policy-relevant in the context of urban poverty.

### 2025-07-03 docs/paper.tex
write data & empirical strategy

>>>>

\section{Data and Empirical Strategy}

The analysis draws on multiple nationally representative data sources,
with the Philippine Censuses serving as the primary dataset. Microdata
from the 1990, 1995, 2000, 2007, and 2010 waves are used to construct
baseline comparisons and perform robustness checks. The 1995 wave is
treated as the pre-policy reference period, while 2007 serves as the
post-policy period. Due to the possibility that reductions in
contraceptive access began prior to the formal issuance of Executive
Order No.~003 in 2000, the 2000 census wave is excluded from the
baseline specification, though it is incorporated in secondary analyses.
Placebo estimations using the 1990 census are also implemented to
validate the identifying assumptions.

While the census provides rich information on household composition and
allows for city-level fixed effects with considerable statistical power,
it does not include direct measures of child health or anthropometry. To
address this limitation, two additional datasets are employed. The
Philippine Demographic and Health Surveys (DHS), pooled across the 1993,
1998, 2003, and 2008 rounds, offer standardized anthropometric data for
children under five, alongside detailed maternal characteristics and
fertility histories. The DHS enables construction of child-level
outcomes---height-for-age z-scores (HAZ) and weight-for-height z-scores
(WHZ)---and supports temporal comparisons across the pre- and post-ban
periods. These outcomes serve as internationally recognized indicators
of chronic and acute malnutrition, respectively.

To complement these sources, the Annual Poverty Indicators Survey (APIS)
is used to assess household-level economic conditions, particularly food
expenditure, living standards, and maternal education. Although smaller
in scale, the APIS provides contextual variables necessary for exploring
heterogeneity in treatment effects across wealth strata or educational
backgrounds.

The empirical strategy adopts a difference-in-differences (DiD)
framework. It utilizes the 2000 contraceptive policy in Manila as a
natural experiment. The policy---enacted through Executive Order
No.~003---effectively banned modern contraceptive distribution in public
health facilities across Manila City, but left surrounding
municipalities in the National Capital Region (NCR) unaffected. This
jurisdictional variation enables identification of the causal effect of
fertility increases on child nutrition, under the assumption that, in
the absence of the policy, nutritional trends in Manila would have
followed a parallel trajectory to those in comparison cities.

The main estimating equation is specified as follows:

\[
\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict},
\]

where \(\text{Nutrition}_{ict}\) denotes the anthropometric outcome (HAZ
or WHZ) for child \(i\) in city \(c\) at time \(t\); \(\text{Post}_t\)
is an indicator for the post-policy period; \(\text{Manila}_c\) is a
treatment group dummy; \(X_{ict}\) is a vector of controls including
child age, sex, birth order, maternal age and education, household
wealth quintile, and urban barangay indicators; \(\lambda_c\) and
\(\tau_t\) are city and year fixed effects, respectively. The
coefficient \(\beta\) estimates the average treatment effect of the
fertility shock on nutritional outcomes.

To strengthen identification, additional specifications are estimated.
City-specific linear trends are introduced to account for differential
temporal patterns in nutrition outcomes. A triple-difference extension
interacts the DiD term with maternal age cohort indicators, under the
assumption that older women---being less fecund---were less likely to
experience changes in fertility behavior. This approach allows for
control of unobserved city-level shocks that may disproportionately
affect certain demographic groups. Sensitivity analyses also test
alternative control groups: first, using only Quezon City, which is
geographically contiguous and socio-economically comparable to Manila;
and second, using all other NCR cities excluding Quezon.

Standard errors are clustered at the city-year level to allow for serial
correlation and heteroskedasticity. Placebo tests using the 1990--1995
period confirm that pre-policy trends in nutritional outcomes were
parallel across treatment and control groups. Additional robustness
checks include falsification tests using children born outside of the
city jurisdiction and exclusion of households with missing maternal
linkage or implausible anthropometric values.

Although the analysis is subject to the usual limitations of
observational designs---including residual confounding and measurement
error---the combined use of census, DHS, and APIS data enables a
multi-faceted examination of the Q--Q trade-off in nutrition. The
strategy mirrors that of \citet{dumas2019sex}, who investigated
educational outcomes using the same policy shock, but extends their
framework to explore a dimension of child quality that is both
biologically foundational and policy-relevant in the context of urban
poverty.
